A Responsible Machine Learning Workflow\\\normalsize{With Focus on Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2020-02-03 article-information-2019.tex
intermediate commit, checking in-text citations, stoppped at [1], [18], [19]

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}.} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is becoming clear. ML models, a common application of AI, can present serious risks. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{\textit{When a Computer Program Keeps You in Jail}}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}, \cite{shokri2019privacy}. This manuscript makes no claim that these interdependent issues of opaqueness, discrimination, privacy harms, and security vulnerabilities in ML have been resolved, even as singular entities, and much less as complex intersectional phenomena. However, Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures, mostly in the form interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available Python tools, that responsible practitioners can use to address a subset of these vexing problems.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve SR 11-7, and the European Union (EU) Greater Data Privacy Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

@@

The simulated data (see Subsection \ref{ssec:sim_data}) is based on the well-known Friedman datasets. Its known feature importance and discrimination characteristics are used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1979tree}, \cite{friedman1991multivariate}. The mortgage data (see Subsection \ref{ssec:mort_data}) is sourced from the Home Mortgage Disclosure Act (HMDA) database and is a fairly realistic data source for demonstrating the template workflow.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} To provide a sense of accuracy differences, performance of more interpretable constrained ML models and less interpretable unconstrained ML models is compared on simulated data and collected mortgage data. Because the unconstrained ML models, gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), do not exhibit convincing accuracy benefits on the simulated or mortgage data and can also present previously delineated risks, further explanation and discrimination analyses are applied only to the constrained, interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} (See subsection \ref{ssec:mgbm}) and XNNs (see Subsection \ref{ssec:xnn}, \cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analyses. MGBM and XNN interpretable model architectures are selected for the example workflow because they are straightforward variants of popular unconstrained ML models. If practitioners are working with GBM and ANN models, it should be relatively uncomplicated to also evaluate the constrained versions of these models.

@@

The mortgage dataset analyzed here is a random sample of consumer-anonymized loans from the HDMA database. These loans are a subset of all originated mortgage loans in the 2018 HMDA data that were chosen to represent a relatively comparable group of consumer mortgages.  A selection of features is used to predict whether a loan is \textit{high-priced}, i.e., the annual percentage rate (APR) charged was 150 basis points (1.5\%) or more above a survey-based estimate of other similar loans offered around the time of the given loan.  After data cleaning and preprocessing to encode categorical features and create missing markers, the mortgage data contains ten input features and the binary outcome, \textit{high-priced}. The data is split into a training set with 160,338 loans and a marker for 5-fold cross-validation and a test set containing 39,662 loans. While lenders would almost certainly use more information than the selected features to determine whether to offer and originate a high-priced loan, the selected input features (LTV ratio, DTI ratio, property value, loan amount, introductory interest rate, customer income, etc.) are likely to be some of the most influential factors that a lender would consider. See Appendix \ref{a_sec:mort} for general information regarding HMDA data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Monotonic Gradient Boosting Machine}\label{ssec:mgbm}

@@

\subsection{Explainable Neural Network}\label{ssec:xnn}

@@

Shapley explanations are a class of additive, locally accurate feature contribution measures with long-standing theoretical support \cite{shapley}, \cite{shapley1988shapley}. Shapley explanations are the only locally accurate and globally consistent feature contribution values, meaning that Shapley explanation values for input features always sum to $g(\mathbf{x})$ for any $\mathbf{x}$, and that Shapley explanation values should not decrease in magnitude for some $x_j$ when $g$ is changed such that $x_j$ truly makes a stronger contribution to $g(\mathbf{x})$ \cite{shapley}, \cite{tree_shap}. Shapley values can be estimated in different ways, many of which are intractable for datasets with large numbers of input features. Tree SHAP (SHapley Additive exPlanations) is a specific implementation of Shapley explanations that relies on traversing internal decision tree structures to efficiently estimate the contribution of each $x_j$ for some $g(\mathbf{x})$ \cite{tree_shap}. Tree SHAP has been implemented in popular gradient boosting libraries such as \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2ogradientboostingestimator}{\texttt{h2o}}, \href{https://lightgbm.readthedocs.io/en/latest/Python-Intro.html}{\texttt{LightGBM}}, and \href{https://xgboost.readthedocs.io/en/latest/python/python_intro.html}{\texttt{XGBoost}}, and Tree SHAP is used to calculate accurate and consistent global and local feature importance for MGBM models in Section \ref{sec:res} and Appendix \ref{a_sec:res_sim}. Deep SHAP is an approximate Shapley value technique that creates SHAP values for ANNs \cite{shapley}. Deep SHAP is implemented in the \href{https://github.com/slundberg/shap}{\texttt{shap}} package and is used to generate SHAP values for the two $g^{XNN}$ models discussed in Section \ref{sec:res} and Appendix \ref{a_sec:res_sim}. For more information pertaining to the calculation of Shapley values, see Appendices \ref{a_ssec:not} and \ref{a_ssec:shap}.

@@

Because many technical and academic discussions of fairness in ML have been inconclusive\footnote{See: \href{https://www.youtube.com/watch?v=jIXIuYdnyyk}{Tutorial: 21 Fairness Definitions and Their Politics}.}, this text will draw on regulatory and legal standards that have been used for years in regulated, high-stakes employment and financial decisions. The discussed measures are also representative of fair lending analyses and pair well with the mortgage data. See Appendix \ref{a_sec:types} for a brief discussion regarding different types of discrimination under US laws and Appendix \ref{a_sec:prac_v_stat} for remarks on statistical vs. practical significance for discrimination measures. One common measure of disparate impact used in US litigation and regulatory settings is known as marginal effects (ME). ME is simply the difference between the percent of the control group members receiving a favorable outcome and the percent of the protected class members receiving a favorable outcome.

@@

\noindent where $X_p$ and $X_c$ represent binary markers created from some demographic attribute, $c$ denotes the control group (often whites or males), $p$ indicates a protected group, and $\text{Pr}(\cdot)$ is the operator for conditional probability.  ME is a favored DI metric used by the the US Consumer Finance Protection Bureau (CFPB), the primary agency charged with regulating fair lending laws at the largest US lending institutions and for various other participants in the consumer financial market.\footnote{See: \href{https://files.consumerfinance.gov/f/201510_cfpb_supervisory-highlights.pdf}{Supervisory Highlights, Issue 9, Fall 2015}.} Another important DI measure is adverse impact ratio (AIR), more commonly known as a \textit{relative risk ratio} in settings outside of regulatory compliance.

@@

\noindent AIR is equal to the ratio of the proportion of the protected class that receives a favorable outcome divided by the proportion of the control class that receives a favorable outcome. Statistically significant AIR values below 0.8 can be considered \textit{prima facia} evidence of discrimination. An additional long-standing measure of DI is standardized mean difference (SMD). SMD is often used to assess disparities in continuous features, such as income differences in employment analyses, or interest rate differences in lending. It originates from work on statistical power, and is more formally known as \textit{Cohen’s d}. SMD is equal to the difference in the average protected class outcome, $\bar{\hat{\mathbf{y}}}_p$, minus the control class outcome, $\bar{\hat{\mathbf{y}}}_c$, divided by a measure of the standard deviation of the population, $\sigma_{\hat{\mathbf{y}}}$.\footnote{There are several measures of the standard deviation of the score that are typically used: \textbf{1.} the standard deviation of the population, irrespective of protected class status, \textbf{2.} a standard deviation calculated only over the two groups being considered in a particular calculation, or \textbf{3.} a pooled standard deviation, using the standard deviations for each of the two groups with weights.} Cohen defined values of this metric to have \textit{small}, \textit{medium}, and \textit{large} effect sizes if the values exceeded 0.2, 0.5, and 0.8, respectively.

@@

\noindent The numerator in the SMD is equivalent to marginal effects but adds the standard deviation divisor as a standardizing factor.  Because of this standardization factor, SMD allows for a comparison across different types of outcomes, such as inequity in mortgage closing fees or inequities in the interest rates given on certain loans.  In this, one may apply definitions in \citet{cohen1988statistical} of \textit{small}, \textit{medium}, and \textit{large} effect sizes, which represent a measure of \textit{practical significance}, which is described in detail in Appendix \ref{a_sec:prac_v_stat}. Finally, confusion matrix measures in demographic groups, such as accuracy, false negative rate (FNR), and their ratios, are also considered as measures of DI in Section \ref{sec:res} and Appendix \ref{a_sec:res_sim}.

@@

Fit comparisons between unconstrained and constrained models and XNN interpretabililty results are discussed in Subsections \ref{ssec:c_v_uc_sim} and \ref{ssec:xnn_sim}. As model training and assessment on the simulated data is a rough validation exercise meant to showcase expected results on data with known characteristics, and given that most of the techniques in the proposed workflow are already used widely or have been validated elsewhere, reporting of simulated data results in the main text will focus mostly on challenging the perceived \textit{accuracy-interpretability tradeoff} for the considered model architectures and on the more novel XNN interpretability results. The bulk of the post-hoc explanation and discrimination testing results for the simulated data are left to Appendix \ref{a_sec:res_sim}.

@@

% TODO: add error metrics, confirm statements about results
\subsubsection{$g^\text{XNN}$ Interpretability Results}\label{ssec:xnn_sim}

@@

Table \ref{tab:mort_acc} shows that $g^\text{ANN}$ and $g^\text{XNN}$ noticeably outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data. This is at least partially due to the preprocessing required to present directly comparable post-hoc explainability results and to use neural networks and tensorflow, e.g., numerical encoding of categorical features and missing values. This preprocessing appears to hamstring some of the tree-based models' inherent capabilities. $g^\text{GBM}$ models trained on non-encoded data with missing values repeatedly produced AUC values of $\sim0.81$ (not shown, but available in resources discussed in Subsection \ref{ssec:soft}).

@@

\subsubsection{Interpretability and Post-hoc Explanation Results}

For $g^\text{MGBM}$, intrinsic interpretability is evaluated with PD and ICE plots of mostly monotonic prediction behavior for several important $X_j$, and post-hoc Shapley explanation analysis is used to create global and local feature importance. Global Shapley feature importance for $g^\text{MGBM}$ on the mortgage test data is reported in Figure \ref{fig:mort_mgbm_glob}. $g^\text{MGBM}$ places high importance on loan-to-value (LTV) ratio, perhaps too high, and also weighs debt-to-income (DTI) ratio, property value, loan amount, and introductory rate period heavily in many of its predictions. Tree SHAP values are reported in the margin space, prior to the application of the logit link function, and the reported numeric values can be interpreted as the mean absolute impact of each $X_j$ on $g^\text{MGBM}(\mathbf{X})$ in the mortgage test data in the $g^\text{MGBM}$ margin space.

@@

PD and ICE are displayed with a histogram in Section \ref{sec:res} to highlight any sparse regions in an input feature's domain. Because most ML models will always issue a prediction on any datum with a correct schema, it's crucial to consider whether a given model learned enough about an observation to make an accurate prediction. Viewing PD and ICE along with a histogram is a convenient method to visually assess whether a prediction is reasonable and based on sufficient training data. DTI ratio and LTV ratio do appear to have sparse regions in their univariate distributions. The monotonicity constraints likely play to the advantage of $g^\text{MGBM}$ in this regard, as $g^\text{MGBM}$ appears to carry reasonable predictions learned from populous domains into the sparse domains of both features.

@@

Tables \ref{tab:mort_disc}\textbf{a} and \ref{tab:mort_disc}\textbf{b} show the results of the discrimination tests using the mortgage data for two sets of class-control groups: blacks as compared to whites, and females as compared to males. As with the simulated data in Table \ref{tab:sim_disc}, several measures of disparities are shown, with the SMDs calculated using the probabilities from $g^{\text{MGBM}}$ and $g^{\text{XNN}}$, and the accuracy, false positive rates (FPRs), and  FPR ratios, MEs, and AIRs calculated using a binary outcome based on a cutoff of 0.20 (anyone with probabilities of 0.2 or less receives the favorable outcome).\footnote{See Appendix \ref{a_sec:cut} for comments pertaining to cutoff selection and discrimination testing.} Since $g^{\text{MGBM}}$ and $g^{\text{XNN}}$  are predicting the likelihood of receiving a high-priced loan, $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ assume that a lower score is favorable.  Thus, one might consider FPR ratios as a measure of the class-control disparities. FPR ratios are higher under $g^{\text{XNN}}$ than $g^{\text{MGBM}}$ (2.45 vs. 2.10) in Table \ref{tab:mort_disc}\textbf{b}, but overall FPRs are lower for blacks under $g^{\text{XNN}}$ (0.295 vs. 0.315) in Table \ref{tab:mort_disc}\textbf{a}.  This is the same pattern seen in the simulated data results in Section \ref{ssec:dis_sim}, again leading to the question of whether a fairness goal should not only consider class-control relative rates, but also intra-class improvements in the chosen fairness metric.  Similar results are found for the female-male comparison, buat the relative rates are less stark: 1.15 for $g^{\text{MGBM}}$ and 1.21 for $g^{\text{XNN}}$.

>>>>

% - Structure ------------------------- x 
% - External links
% - Internal links
%   - Sections/subsections/appendices - x
%   - References
%   - Tables/figures
%   - footnotes
% - Abbreviations --------------------- x
% - Bibliography 
%   - Nick/Bryce citation
% - Notation
% - Tables/figures
% - Grammar and spelling
%   - Hyphenation

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}.} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is becoming clear. ML models, a common application of AI, can present serious risks. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{\textit{When a Computer Program Keeps You in Jail}}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}, \cite{shokri2019privacy}. This manuscript makes no claim that these interdependent issues of opaqueness, discrimination, privacy harms, and security vulnerabilities in ML have been resolved, even as singular entities, and much less as complex intersectional phenomena. However, Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures, mostly in the form interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available Python tools, that responsible practitioners can use to address a subset of these vexing problems.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve SR 11-7, and the European Union (EU) General Data Protection Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

@@

The simulated data (see Subsection \ref{ssec:sim_data}) is based on the well-known Friedman datasets. Its known feature importance and discrimination characteristics are used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1979tree}, \cite{friedman1991multivariate}. The mortgage data (see Subsection \ref{ssec:mort_data}) is sourced from the Home Mortgage Disclosure Act (HMDA) database and is a fairly realistic data source for demonstrating the template workflow.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} To provide a sense of accuracy differences, performance of more interpretable constrained ML models and less interpretable unconstrained ML models is compared on simulated data and collected mortgage data. Because the unconstrained ML models, gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), do not exhibit convincing accuracy benefits on the simulated or mortgage data and can also present unmitigated risks, further explanation and discrimination analyses are applied only to the constrained, interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} (See Subsection \ref{ssec:mgbm}) and XNNs (see Subsection \ref{ssec:xnn}, \cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analyses. MGBM and XNN interpretable model architectures are selected for the example workflow because they are straightforward variants of popular unconstrained ML models. If practitioners are working with GBM and ANN models, it should be relatively uncomplicated to also evaluate the constrained versions of these models.

@@

The mortgage dataset analyzed here is a random sample of consumer-anonymized loans from the HDMA database. These loans are a subset of all originated mortgage loans in the 2018 HMDA data that were chosen to represent a relatively comparable group of consumer mortgages.  A selection of features is used to predict whether a loan is \textit{high-priced}, i.e., the annual percentage rate (APR) charged was 150 basis points (1.5\%) or more above a survey-based estimate of other similar loans offered around the time of the given loan.  After data cleaning and preprocessing to encode categorical features and create missing markers, the mortgage data contains ten input features and the binary outcome, \textit{high-priced}. The data is split into a training set with 160,338 loans and a marker for 5-fold cross-validation and a test set containing 39,662 loans. While lenders would almost certainly use more information than the selected features to determine whether to offer and originate a high-priced loan, the selected input features (loan to value (LTV) ratio, debt to income (DTI) ratio, property value, loan amount, introductory interest rate, customer income, etc.) are likely to be some of the most influential factors that a lender would consider. See Appendix \ref{a_sec:mort} for general information regarding HMDA data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Monotonic Gradient Boosting Machines}\label{ssec:mgbm}

@@

\subsection{Explainable Neural Networks}\label{ssec:xnn}

@@

Shapley explanations are a class of additive, locally accurate feature contribution measures with long-standing theoretical support \cite{shapley}, \cite{shapley1988shapley}. Shapley explanations are the only locally accurate and globally consistent feature contribution values, meaning that Shapley explanation values for input features always sum to $g(\mathbf{x})$ for any $\mathbf{x}$, and that Shapley explanation values should not decrease in magnitude for some $x_j$ when $g$ is changed such that $x_j$ truly makes a stronger contribution to $g(\mathbf{x})$ \cite{shapley}, \cite{tree_shap}. Shapley values can be estimated in different ways, many of which are intractable for datasets with large numbers of input features. Tree SHAP (SHapley Additive exPlanations) is a specific implementation of Shapley explanations that relies on traversing internal decision tree structures to efficiently estimate the contribution of each $x_j$ for some $g(\mathbf{x})$ \cite{tree_shap}. Tree SHAP has been implemented in popular gradient boosting libraries such as \href{http://docs.h2o.ai/h2o/latest-stable/h2o-py/docs/modeling.html#h2ogradientboostingestimator}{\texttt{h2o}}, \href{https://lightgbm.readthedocs.io/en/latest/Python-Intro.html}{\texttt{LightGBM}}, and \href{https://xgboost.readthedocs.io/en/latest/python/python_intro.html}{\texttt{XGBoost}}, and Tree SHAP is used to calculate accurate and consistent global and local feature importance for MGBM models in Subsection \ref{ssec:int_ex_mort} and Appendix \ref{a_ssec:int_ex_sim}. Deep SHAP is an approximate Shapley value technique that creates SHAP values for ANNs \cite{shapley}. Deep SHAP is implemented in the \href{https://github.com/slundberg/shap}{\texttt{shap}} package and is used to generate SHAP values for the two $g^{XNN}$ models discussed in Subsection \ref{ssec:int_ex_mort} and Appendix \ref{a_ssec:int_ex_sim}. For more information pertaining to the calculation of Shapley values, see Appendices \ref{a_ssec:not} and \ref{a_ssec:shap}.

@@

Because many technical and academic discussions of fairness in ML have been inconclusive\footnote{See: \href{https://www.youtube.com/watch?v=jIXIuYdnyyk}{Tutorial: 21 Fairness Definitions and Their Politics}.}, this text will draw on regulatory and legal standards that have been used for years in regulated, high-stakes employment and financial decisions. The discussed measures are also representative of fair lending analyses and pair well with the mortgage data. See Appendix \ref{a_sec:types} for a brief discussion regarding different types of discrimination under US laws and Appendix \ref{a_sec:prac_v_stat} for remarks on practical vs. statistical significance for discrimination measures. One common measure of DI used in US litigation and regulatory settings is ME. ME is simply the difference between the percent of the control group members receiving a favorable outcome and the percent of the protected class members receiving a favorable outcome.

@@

\noindent where $X_p$ and $X_c$ represent binary markers created from some demographic attribute, $c$ denotes the control group (often whites or males), $p$ indicates a protected group, and $\text{Pr}(\cdot)$ is the operator for conditional probability.  ME is a favored DI metric used by the the US Consumer Financial Protection Bureau (CFPB), the primary agency charged with regulating fair lending laws at the largest US lending institutions and for various other participants in the consumer financial market.\footnote{See: \href{https://files.consumerfinance.gov/f/201510_cfpb_supervisory-highlights.pdf}{Supervisory Highlights, Issue 9, Fall 2015}.} Another important DI measure is AIR, more commonly known as a \textit{relative risk ratio} in settings outside of regulatory compliance.

@@

\noindent AIR is equal to the ratio of the proportion of the protected class that receives a favorable outcome divided by the proportion of the control class that receives a favorable outcome. Statistically significant AIR values below 0.8 can be considered \textit{prima facia} evidence of discrimination. An additional long-standing measure of DI is SMD. SMD is often used to assess disparities in continuous features, such as income differences in employment analyses, or interest rate differences in lending. It originates from work on statistical power, and is more formally known as \textit{Cohen’s d}. SMD is equal to the difference in the average protected class outcome, $\bar{\hat{\mathbf{y}}}_p$, minus the control class outcome, $\bar{\hat{\mathbf{y}}}_c$, divided by a measure of the standard deviation of the population, $\sigma_{\hat{\mathbf{y}}}$.\footnote{There are several measures of the standard deviation of the score that are typically used: \textbf{1.} the standard deviation of the population, irrespective of protected class status, \textbf{2.} a standard deviation calculated only over the two groups being considered in a particular calculation, or \textbf{3.} a pooled standard deviation, using the standard deviations for each of the two groups with weights.} Cohen defined values of this metric to have \textit{small}, \textit{medium}, and \textit{large} effect sizes if the values exceeded 0.2, 0.5, and 0.8, respectively.

@@

\noindent The numerator in the SMD is equivalent to ME but adds the standard deviation divisor as a standardizing factor.  Because of this standardization factor, SMD allows for a comparison across different types of outcomes, such as inequity in mortgage closing fees or inequities in the interest rates given on certain loans.  In this, one may apply definitions in \citet{cohen1988statistical} of \textit{small}, \textit{medium}, and \textit{large} effect sizes, which represent a measure of \textit{practical significance}, which is described in detail in Appendix \ref{a_sec:prac_v_stat}. Finally, confusion matrix measures in demographic groups, such as accuracy, false positive rate (FPR), and their ratios, are also considered as measures of DI in Subsection \ref{ssec:dis_mort} and Appendix \ref{a_ssec:dis_sim}.

@@

Fit comparisons between unconstrained and constrained models and XNN interpretabililty results are discussed in Subsections \ref{ssec:c_v_uc_sim} and \ref{ssec:xnn_sim}. As model training and assessment on the simulated data is a rough validation exercise meant to showcase expected results on data with known characteristics, and given that most of the techniques in the proposed workflow are already used widely or have been validated elsewhere, reporting of simulated data results in the main text will focus mostly fit measures for the considered data and model architectures and on the more novel $g^\text{XNN}$ interpretability results. The bulk of the post-hoc explanation and discrimination testing results for the simulated data are left to Appendix \ref{a_sec:res_sim}.

@@

% TODO: add error metrics, confirm statements about results
\subsubsection{Interpretability Results}\label{ssec:xnn_sim}

@@

Table \ref{tab:mort_acc} shows that $g^\text{ANN}$ and $g^\text{XNN}$ noticeably outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data. This is at least partially due to the preprocessing required to present directly comparable post-hoc explainability results and to use neural networks and tensorflow, e.g., numerical encoding of categorical features and missing values. This preprocessing appears to hamstring some of the tree-based models' inherent capabilities. $g^\text{GBM}$ models trained on non-encoded data with missing values repeatedly produced area under the curve (AUC) values of $\sim0.81$ (not shown, but available in resources discussed in Subsection \ref{ssec:soft}).

@@

\subsubsection{Interpretability and Post-hoc Explanation Results}\label{ssec:int_ex_mort}

For $g^\text{MGBM}$, intrinsic interpretability is evaluated with PD and ICE plots of mostly monotonic prediction behavior for several important $X_j$, and post-hoc Shapley explanation analysis is used to create global and local feature importance. Global Shapley feature importance for $g^\text{MGBM}$ on the mortgage test data is reported in Figure \ref{fig:mort_mgbm_glob}. $g^\text{MGBM}$ places high importance on LTV ratio, perhaps too high, and also weighs DTI ratio, property value, loan amount, and introductory rate period heavily in many of its predictions. Tree SHAP values are reported in the margin space, prior to the application of the logit link function, and the reported numeric values can be interpreted as the mean absolute impact of each $X_j$ on $g^\text{MGBM}(\mathbf{X})$ in the mortgage test data in the $g^\text{MGBM}$ margin space.

@@

PD and ICE are displayed with a histogram to highlight any sparse regions in an input feature's domain. Because most ML models will always issue a prediction on any datum with a correct schema, it's crucial to consider whether a given model learned enough about an observation to make an accurate prediction. Viewing PD and ICE along with a histogram is a convenient method to visually assess whether a prediction is reasonable and based on sufficient training data. DTI ratio and LTV ratio do appear to have sparse regions in their univariate distributions. The monotonicity constraints likely play to the advantage of $g^\text{MGBM}$ in this regard, as $g^\text{MGBM}$ appears to carry reasonable predictions learned from populous domains into the sparse domains of both features.

@@

Tables \ref{tab:mort_disc}\textbf{a} and \ref{tab:mort_disc}\textbf{b} show the results of the discrimination tests using the mortgage data for two sets of class-control groups: blacks as compared to whites, and females as compared to males. As with the simulated data in Table \ref{tab:sim_disc}, several measures of disparities are shown, with the SMDs calculated using the probabilities from $g^{\text{MGBM}}$ and $g^{\text{XNN}}$, and the accuracy, FPRs, and  FPR ratios, MEs, and AIRs calculated using a binary outcome based on a cutoff of 0.20 (anyone with probabilities of 0.2 or less receives the favorable outcome).\footnote{See Appendix \ref{a_sec:cut} for comments pertaining to discrimination testing and cutoff selection.} Since $g^{\text{MGBM}}$ and $g^{\text{XNN}}$  are predicting the likelihood of receiving a high-priced loan, $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ assume that a lower score is favorable.  Thus, one might consider FPR ratios as a measure of the class-control disparities. FPR ratios are higher under $g^{\text{XNN}}$ than $g^{\text{MGBM}}$ (2.45 vs. 2.10) in Table \ref{tab:mort_disc}\textbf{b}, but overall FPRs are lower for blacks under $g^{\text{XNN}}$ (0.295 vs. 0.315) in Table \ref{tab:mort_disc}\textbf{a}.  This is the same pattern seen in the simulated data results in Appendix \ref{a_ssec:dis_sim}, again leading to the question of whether a fairness goal should not only consider class-control relative rates, but also intra-class improvements in the chosen fairness metric.  Similar results are found for the female-male comparison, buat the relative rates are less stark: 1.15 for $g^{\text{MGBM}}$ and 1.21 for $g^{\text{XNN}}$.
